---
title: "Health and Economic Impact of Severe Weather Events"
author: "J Faleiro"
date: "March 21st, 2016"
output: html_document
---

### Synopsis

The basic goal of this study is to explore the NOAA Storm Database and answer some basic questions about health and economic impact of severe weather events across the United States, specifically:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

Our analysis will be conducted in R and will consist of tables, figures and several summaries. We will take special care to document each of the steps, making sure they follow a fully scripted flow and therefore allow for the entire analysis to be fully reproducible.

### Data Processing

This section describes and executes the preparation of the data required for our analysis, from dowloading the data, to loading the data into R, to the extraction of samples of the data. Given the size of the data and extensive plots, we will be caching each of the chunks to minimize re-work and speed up processing.

#### Preliminaries

We will start by defining the libraries this analysis depends on. We will define dependencies using `pacman`.

```{r imports, warning=FALSE, message=FALSE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(R.utils, data.table, R.cache, dplyr, tidyr, ggplot2)
```

If dependencies are not in your environment they will be installed and properly loaded.

#### Loading Storm Database

Next, we download the original NOAA storm database to local disk. The location of the remote file is given by the variable `url`.


```{r download, cache=TRUE}
stormFileName <- 'stormdata.csv.bz2'
url <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
download.file(url, stormFileName, method='curl')
```

The file is a comma separated value, compressed.

```{r load_storm_file, cache=TRUE, dependson='download'}
storm <- read.csv(stormFileName, stringsAsFactors=FALSE)
```

We don't know exactly which columns will be needed for this analysis, so we will proceed with a quick inspection of the storm data to understand which columns we will be needed.

```{r}
head(storm)
```

As we can see, this is an extensive dataset. We will need just a subset of the original columns to answer the questions posed in the synopis, specifically:

* The type of event `EVTYPE`
* The state `STATE`
* Number of fatalities `FATALITIES`
* Number of injuries `INJURIES`

```{r extract_columns, dependson='load_storm_file'}
storm <- storm %>% 
         mutate(EVTYPE=toupper(EVTYPE)) %>%
         select(EVTYPE, STATE, FATALITIES, INJURIES)
summary(storm)
```

This analysis will have to be performed at both federal and state level, so we will prepare an aggregate of all fatalitites and injuries across all states. But before that, one last check for absent values:

```{r}
sum(is.na(storm))
```

No absent values.


```{r storm_us_before, dependson='extract_columns'}
stormUS <- storm %>% 
           select(EVTYPE, FATALITIES, INJURIES) %>%
           group_by(EVTYPE) %>%
           summarize(FATALITIES=sum(FATALITIES), INJURIES=sum(INJURIES))
summary(stormUS)
```

#### Adjusting Event Types

Next we will make sure event types, given by column `EVTYPE`, are not misrepresenting events. In other words, we don't have mispelled descriptions. Since we are only concerned with high impact events, we will only check the top events.


```{r}
sampleSize <- 15
head(arrange(stormUS, desc(FATALITIES)), sampleSize)
```

There are several event types described by the different `EVTYPE` values. We will proceed with cleaning up of the more important ones. We will use `dplyr::mutate` and `grepl` to look for similar event types and replace them by a representative alternative:

```{r storm, dependson='extract_columns'}
storm <- storm %>%
         mutate(EVTYPE=ifelse(grepl('.*HEAT', EVTYPE), 'HEAT', 
                       ifelse(grepl('.*WIND', EVTYPE), 'STRONG WIND',
                       ifelse(grepl('.*TORNADO.*', EVTYPE), 'TORNADO',
                       ifelse(grepl('.*HURRICANE.*', EVTYPE), 'HURRICANE',
                       ifelse(grepl('.*SURF.*', EVTYPE), 'HIGH SURF',
                       ifelse(grepl('.*RIP CURRENT.*', EVTYPE), 'RIP CURRENT',
                       ifelse(grepl('.*FLASH FLOOD.*', EVTYPE), 'FLASH FLOOD',
                       ifelse(grepl('.*UNSEASONABLY.*WARM.*', EVTYPE), 'UNSEASOBNABLY WARM',
                       ifelse(grepl('.*UNSEASONABLY.*COLD.*', EVTYPE), 'UNSEASOBNABLY COLD',
                       ifelse(grepl('.*WINTER.*', EVTYPE), 'WINTER WEATHER',
                       ifelse(grepl('.*WILD.*FIRE.*', EVTYPE), 'WILD FIRE',
                              EVTYPE)))))))))))
                )
```


And since we changed several important event types, we will re-do the U.S. aggregation of events and re-sample it:


```{r storm_us, dependson='storm'}
stormUS <- storm %>% 
           select(EVTYPE, FATALITIES, INJURIES) %>%
           group_by(EVTYPE) %>%
           summarize(FATALITIES=sum(FATALITIES), INJURIES=sum(INJURIES))

head(arrange(stormUS, desc(FATALITIES)), sampleSize)
```

Now the more important events have a representative and unique event type. Much better.

#### Scaling Quantities

In order to come up with a notion of _impact_ we need quantities of fatalities to be somehow comparable. We wil be scaling them within a minimum and maximum interval so for every quantity $q$ it will be bound to the interval $0 <= q <= 1$. We will use the function `doScaling` for that.

```{r}
doScaling <- function(items) {
    # scale(items, center=FALSE, scale=max(items)) # does not work properly 
    if (min(items) == max(items)) 1 else (items - min(items)) / (max(items) - min(items))
}
```

For a meaningfull representation of fatalities and injuries, we have to scale them separately:
 
```{r storm_scaled, dependson='storm'}
storm$FATALITIES <- doScaling(storm$FATALITIES)
storm$INJURIES <- doScaling(storm$INJURIES)
summary(storm)
```

We have to perform the same scaling on data across U.S. as well, again separately for fatalities and injuries.

```{r storm_us_scaled, dependson='storm_us'}
stormUS$FATALITIES <- doScaling(stormUS$FATALITIES)
stormUS$INJURIES <- doScaling(stormUS$INJURIES)
summary(stormUS)
```

#### De-Leveraging

We can see in the summary above the high leverage of values on `FATALITIES` and `INJURIES`. Given the nature of the questions we intend to answer, we care about high leveraged quantities only. We will de-leverage data at U.S. and state levels separately.

##### Removing U.S. Low Leverage Data

```{r deleveraged-us, dependson='storm_us_scaled'}
top <- 7
topFatalities <- sort(stormUS$FATALITIES, TRUE)[top] # top fatalities
topInjuries <- sort(stormUS$INJURIES, TRUE)[top]     # top injuries
majorStormsUS <- stormUS %>%
                 filter(FATALITIES > 0 & INJURIES > 0) %>%
                 filter(FATALITIES >= topFatalities & INJURIES >= topInjuries)
```

##### Removing State Low Leverage Data

```{r deleveraged-state, dependson='storm-scaled'}
top <- 3
majorStormsInjuries <- storm %>%
                            select(EVTYPE, STATE, INJURIES) %>%
                            filter(INJURIES > 0) %>%
                            group_by(STATE) %>%
                            top_n(n=top, wt=INJURIES)
majorStormsFatalities <- storm %>%
                            select(EVTYPE, STATE, FATALITIES) %>%
                            filter(FATALITIES > 0) %>%
                            group_by(STATE) %>%
                            top_n(n=top, wt=FATALITIES)
```

#### Impact Calculation 

##### Assumptions 

To answer the 2 questions posted on our synopsis, we will first have to differentiate between events that are more harmful to population health and the ones that have the greatest economic consequences. Since not all the data necessary for this analysis is available (like life expectancy for proper calculation of mortality economic cost) we will use a simplification of the [WHO quide to identifying the economic consequences of disease and injury][WHO_guide_economic_impact]:

* Number of injuries will define the economic impact of an event
* Number of fatalities will define the harm impact of an event to population health 

To adequate to this model, we reshape the dataframe to represent economic and health impact, given the following structure:

* The type of event `EVTYPE`
* The state `STATE`
* The occurrence `OCURRENCE`, one of _ECONOMIC_ or _HEALTH_
* The impact value `IMPACT`

The value of impact will quantify economic and health impact at both state and federal levels.

##### Calculation of Impact at State Level

The impact value at the state level is given by datasets `topStormsEconomicImpact` and `topStormsHealthImpact` for economic and health impact respectively.

```{r impact-state, dependson='deleveraged-state'}
topStormsEconomicImpact <- majorStormsInjuries %>%
                           mutate(IMPACT=INJURIES) %>%
                           select(EVTYPE, STATE, IMPACT)
topStormsHealthImpact <- majorStormsFatalities %>%
                           mutate(IMPACT=FATALITIES) %>%
                           select(EVTYPE, STATE, IMPACT)
```

##### Calculation of Impact at Federal Level

The impact value at the federal level is given by dataset `stormImpactUS` for both economic and health impacts.

```{r impact-us, dependson='deleveraged-us'}
stormImpactUS <- majorStormsUS %>%
                 mutate(ECONOMIC=FATALITIES, HEALTH=INJURIES) %>%
                 select(EVTYPE, ECONOMIC, HEALTH) %>%
                 gather(OCCURRENCE, IMPACT, ECONOMIC:HEALTH)
```

### Results

The results of this analysis will be derived from the analysis of three plots: one presents answers to the originals questions at the federal level, and two will approach the same questions at the state level. The questions are repeated here for convenience.

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

We are limited to a maximum of three plots total.

#### Entire U.S. Analysis

The analysis at the U.S. level uses a bar chart plotted through `ggplot2` so show side by side the events that bring the most impact to the U.S. economy and polulation health.

```{r plots-us, dependson='deleveraged-us'}
ggplot(stormImpactUS, aes(x=EVTYPE, y=IMPACT, group=OCCURRENCE, fill=OCCURRENCE, width=.6)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_bar(stat = "identity", position = "dodge", colour = "black", 
           show.legend = FALSE) +
  labs(x = NULL, y = "Impact") +
  ggtitle("Health and Economic Impact of Weather Events - U.S.") +
  theme_bw() +
  theme(panel.grid.major.y = element_line(colour = "grey60"),
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        plot.title = element_text(size = rel(1.2), face = "bold"),
        legend.position = "top",
        legend.title = element_blank())
```

The events of greatest economic and health impact are tornadoes, heat, strong winds, lightning and floods, in descreasing order:

* Answer to question 1: tornadoes, strong wind, heat, flood and lightning
* Answer to question 2: tornadoes, heat, strong wind, lightining and flood.

#### State by State Analysis

Given the large number of states and events, we opted for heatmaps for separate analysis of economic and health impacts.

##### Population health impact

The health impact is given by a heat map in which impact is given by a scale of red.

```{r plot-health-states, fig.width=10, fig.height=10, dependson='deleveraged-state'}
baseSize <- 9
ggplot(topStormsHealthImpact, aes(y=STATE, x=EVTYPE)) + 
    ggtitle('Health Impact of Weather Events - per State') +
    geom_tile(aes(fill=IMPACT), colour="white") + 
    scale_fill_gradient(low="white", high="red", name='Health\nImpact') +
    theme_grey(base_size=baseSize) +
    labs(x='', y='') +
    scale_y_discrete(expand=c(0,0)) +
    theme(axis.text.x=element_text(size=baseSize, angle=270, hjust=0, colour='grey50'))
```

This heat map shows the most prevalent types of high economic impact events: tornadoes, heat, strong wind and flash flood, - and each of the states in which they are more prevalent.

* Answer to question 1 (state level): tornadoes, heat, strong wind and flash flood, affecting different states at different intensities - for more details refer to the heat map above.

##### Economic Impact

The economic impact is given by a heat map in which impact is given by a scale of green.

```{r plot-economic-states, fig.width=10, fig.height=10, dependson='deleveraged-state'}
ggplot(topStormsEconomicImpact, aes(y=STATE, x=EVTYPE)) + 
    ggtitle('Economic Impact of Weather Events - per State') +
    geom_tile(aes(fill=IMPACT), colour="white") + 
    scale_fill_gradient(low="white", high="springgreen4", name='Economic\nImpact') +
    theme_grey(base_size=baseSize) +
    labs(x='', y='') +
    scale_y_discrete(expand=c(0,0)) +
    theme(axis.text.x=element_text(size=baseSize, angle=270, hjust=0, colour='grey50'))
```

This heat map shows the most prevalent types of high economic impact events: tornadoes, ice storms, heat and lightning - and each of the states in which they are more prevalent.

* Answer to question 2 (state level): tornadoes, ice storms, heat and lighting, affecting different states at different intensities - for more details refer to the heat map above.

### Session Configuration Details

```{r}
sessionInfo()
```

[WHO_guide_economic_impact]: http://www.who.int/choice/publications/d_economic_impact_guide.pdf "WHO quide to identifying the economic consequences of disease and injury"
